---
title: "A RAPM Model for EPL Player's Ratings"
author: "Phong Hoang, Edvin Tran Hoac, Boyuan (Gary) Zhang"
date: '2022-07-25'
output: 
  html_document
---

```{=html}
<style type="text/css">

h1.title {
  font-size: 48px;
  color: DarkRed;
  text-align: center;
}
h4.author { 
    font-size: 20px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { 
  font-size: 20px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# **I. Introduction**

Soccer, more commonly known as Football, is a family of team sports that includes the involvement of kicking the ball and scoring the ball. Originated in England in 1863, soccer has gradually become the most popular sport in the world due to its simplicity. This beautiful game can be played from anywhere with minimal equipment. With the improvement of the media, we as soccer fans shouldn't be too unfamiliar with the story of famous soccer players who grew up playing barefoot on the street.

With its popularity, soccer can be viewed as a large global business in which at the highest level, billions of dollars are invested to compete for championships. This enormous size has attracted numerous media, making the process of evaluating individual players much more appealing and resulting in significant impact on the recruitment decisions and drafting decisions. Analysts and researchers have put up hours to construct a framework that is able to deliver one-number statistics ratings for each player, allowing organizations to quickly evaluate trades, pick up free agents, and predict future matches' results.

One of the most well-known frameworks for constructing all-in-one player performance metrics is the plus-minus model, which in the most rudimentary form has been applied in hockey since the 1950s. The plus-minus model considers the number of goals scored minus the number of goals conceded when a given player is in the game. A huge problem with this approach is that it does not control for the impact of teammates or opponents. It is important to acknowledge that every player on the pitch, either directly or indirectly, is contributing to the overall team's performance. Several academic studies have started to utilize linear regression as an adjusted plus-minus (APM) framework to include other players' influence on that individual's rating. APM and its variations have most commonly been seen in basketball and hockey, achieving substantial improvements in these fields, an example being ESPN's widely known real plus-minus (RPM).

As a sport, soccer has numerous inherent disadvantages when it comes to APM, especially compared to basketball or hockey. Soccer is a low-scoring game with few substitutions, which means a traditional APM for the sport will have collinearity issues and an infrequent response variable. The collinearity comes from the low number of substitutions since some players will share the same minutes on the court together in almost every segment, which eventually makes them indistinguishable. Out of these three sports, basketball is the best sport to calculate APM for, and whereas hockey is low-scoring, it has an extremely high number of substitutions every game. Several scholars have tried to handle this challenge, considerably the paper from the Department of Statistics at Carnegie Mellon University, which introduces the use of video game ratings from FIFA as a prior in the APM model.

This paper aims to build up on the foundations of calculating individual player ratings using a plus-minus framework. This procedure ensures that the one-number statistics for soccer players accurately represent the individuals' skill level as well as their team contribution by adding the traditional box-score rating into the measurement of the APM model. Our approach also uses expected goals instead of the actual goals as we believe this will better measure the team's performance within a match. The remainder of this paper will be organized as follows. Section 2 describes the dataset that we use. Section 3 goes into detail about different stages of our method. Sections 4 and 5 summarize the results and discuss the project's limitations and propose several next steps for further research. The last section will be the acknowledgments for people that have helped us to publish this work.

# **II. Dataset**

## **1. Prior Stages**

In the construction of individuals' ratings, we use two different datasets. First of all, we collected the box-score ratings for each player in the Premier League season 2020-21 and 2021-22 from FBref.com, which is the soccer section of the ["Sports Reference Website"](https://www.sports-reference.com/). This data set contains all the players from five majors League in Europe, with each observations being a player along with over 180 variables describing his information and box-score statistics. The variables that are considered describe different actions in soccer, such as scoring, creating, dribbling, passing and defensive actions. We also consider players from EPL and had at least 900 minutes in the field last season in order to reduces the bias in which players that usually start from bench benefits from the team's results.

The second data set we collected is the FIFA ratings from 2021 on the website ["https://sofifa.com/"](https://sofifa.com/). The website contains different characteristics to measure the players, but we stick with the one-number overall statistics for each player only. We merge these two dataset by players' names, resulting in several missing values as the names are recorded slightly different between two data sources. These missingness are handled by manually matching. Figure 1 shows the distribution of EPL FIFA 2022 ratings.

<center>

```{r}
library(ggplot2)
fifarating <- read.csv("Final Project/data/eng2021_boxscores.csv")
ggplot(data=fifarating, aes(x=overall)) +
    geom_histogram(color="lightblue", fill="steelblue", bins = 15) + 
    labs(title="EPL FIFA Rating Season 2022", x = ("FIFA Rating")) + 
  theme(plot.title = element_text(color="Black", size=12, face="bold", hjust = 0.5),
axis.title.x = element_text(color="black", size=12, face="bold"),
axis.title.y = element_text(color="black", size=12, face="bold"))
```
</center>

The distribution is the frequency of the FIFA Rating among players, whereas the x-axis represents the ratings, and the y-axis is the frequency. The histogram is close to a normal distribution, in which a great proportion of the observations is in the range from 72 to 83.

## **2. Modeling Stages**

In this stage, with the purpose of getting the expected goals for each stint, we also have two different dataset. In order to create the stint dataset, we collected the match summary and line-up information for every match in the 2021-2022 English Premier League season. We then defined each separated segment as there is a change in the scoreboard or in the lineup from either team. Therefore, a new stint will begin when there is an event of a substitution, red card, or scoring a goal. This dataset contains almost 4000 stints across 380 matches, which makes a match has an average of over 10 stints. The following plots describes the distribution of number of stints per match in the EPL 2021-22.

<center>

```{r}
library(ggplot2)
library(dplyr)
xG_df <- read.csv("eng1_2122_singleapm_stint_eg (1).csv")
xG_df <- xG_df %>% group_by(matchID) %>% dplyr::summarise(count(Gender))
ggplot(data=xG_df, aes(x=freq)) +
    geom_histogram(color="lightblue", fill="steelblue", bins = 10) + 
    labs(title="Distribution of Stints Per Game EPL 2021-22", x = ("Number of Stints")) + 
  theme(plot.title = element_text(color="Black", size=12, face="bold", hjust = 0.5),
axis.title.x = element_text(color="black", size=12, face="bold"),
axis.title.y = element_text(color="black", size=12, face="bold"))
```
</center>

As we can see from the distributions, almost 150 matches have 9 or 9 stints, while more than 150 teams have the number of stints ranging from 10 to 14. There are several potential outliers which may be resulted from a no-subtition match or multiple goals are scored from both sides.

The next step in this stage is to have the shooting information match-by-match. Once again, we collect this information on ["Sports Reference Website"](https://www.sports-reference.com/) An observation in this data frame represented a shot made by a player in that game, along with its expected goal. We also consider several important factors from this dataset such as the minute of the shot, player's name, team's name, and many more. 